### 2025-01-20 products/manuscript/Manuscript.qmd
Cloned the repo

---
title: "Example Manuscript Template for a Data Analysis Project"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

**Authors**

-   Guozheng Yang$^{1, *}$ (ORCID: 0000-0001-7361-1847)

**Author affiliations**

1.  College of Public Health, University of Georgia, Athens, GA, USA.

$*$ Corresponding author: guozheng0224\@uga.edu

{{< pagebreak >}}

# Part 1

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

## General Background Information

*Provide enough background on your topic that others can understand the why and how of your analysis*

## Description of data and data source

*Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*

## Questions/Hypotheses to be addressed

*State the research questions you plan to answer with this analysis.*

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*

## Data import and cleaning

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-01-31 products/manuscript/Manuscript.qmd
Part 1

@@
 ---
-title: "Example Manuscript Template for a Data Analysis Project"
+title: "Sex-Based Differences in Tuberculosis Immune Response: The Role of Leptin"
 author: ""
@@
 
-# Part 1
+# Abstract
 
-# Summary/Abstract
+## Background:
+
+## Methods:
+
+## Results:
+
+## Conclusions:
+
+# Keywords:
 
@@
 
-*Provide enough background on your topic that others can understand the why and how of your analysis*
+As a global threat, tuberculosis (TB) caused 10.8 million new cases in 2023 with Africa contributing 24% of these cases [@WHO2024].
+Countries with high annual incidence rates are mostly in Africa, where TB incidence is significantly higher than in other regions.
+TB prevention and control strategies have shown progress in recent years.
+However, further efforts are needed to eliminate TB infection and transmission.
 
-## Description of data and data source
+Males are more likely to develop TB disease than females, a difference that cannot be fully explained by socioeconomic status or access to healthcare [@Horton2016; @Ledesma2022].
+Research has shown that this sexual bias is linked to differences in immune systems, influenced by chromosome-encoded genes and hormones [@Gupta2022].
+Previous studies indicate that body fat mass could also affect immune responses [@Womack2007; @Ilavska2012].
+Since women generally have higher body fat storage, their immune systems may receive sufficient energy to support a stronger response to infections compared to men.
+In diseases like TB that often cause body wasting [@Mupere2014], differences in energy reserves may contribute to variations in immune response between men and women.
 
-*Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*
+Leptin, a key link between metabolism and the immune system, plays an important role in activating and regulating immune responses [@Naylor2016; @Francisco2018].
+It functions as an energy indicator, signaling whether the body has enough energy to activate and sustain immune responses [@Lord2002; @Cava2004].
+For individuals with similar body fat mass, females generally have higher serum leptin levels than males [@Hickey1996].
+However, leptin levels may be lower in tuberculosis patients [@Buyukoglan2007], suggesting a suppressed role in regulating metabolism and immune system stability.
+Since women typically have higher body fat stores and leptin levels, the impact on women may be less pronounced compared to men.
+However, there is a lack of evidence to demonstrate this sex difference.
 
-## Questions/Hypotheses to be addressed
-
-*State the research questions you plan to answer with this analysis.*
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+This paper presents findings on sexual differences in metabolism and immune responses in individuals initially infected with TB.
+The primary hypothesis is that women have a stronger immune response to TB due to higher body fat storage.
+Additionally, leptin is hypothesized to mediate immune response based on body energy storage.
 
@@
 # Methods
+## Study design and participants
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
+From March to April 2017, a cross-sectional study was conducted in Kampala, Uganda.
+Sixty participants were recruited from Mulago Hospital, including 30 males and 30 females aged 15 years or older.
+All participants had a first episode of TB confirmed by culture or molecular methods.
+To prevent interference with metabolism and immune responses, individuals with co-morbidities such as asthma, kidney disease, liver disease, cancer, HIV, or diabetes were not enrolled.
+Those receiving hormonal or immune therapies were also excluded.
 
-## Schematic of workflow
+## Body index measures
 
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
+At enrollment, basic body indices, including age, sex, height (m), and weight (kg), were measured.
+Participants who provided informed consent underwent total body water (TBW, L) estimation using deuterium dilution.
+Specifically, participants received a precisely weighed oral dose of deuterium oxide (~25-30g $D_2O$, 99.8% purity) after an overnight fast.
+Participants will be asked to refrain from eating and to consume only the minimum amount of water needed to quench thirst during the 4-hour equilibration period.
+Saliva samples collected before and after a 4-hour equilibration period were analyzed using Fourier transform infrared spectrophotometry to determine deuterium concentration, and TBW was estimated by extrapolating deuterium dilution space, adjusted by a factor of 1.041 [@IAEA2011].
+Based on TBW estimates, lean body mass (LBM, kg) was further estimated assuming a hydration fraction of 0.732 [@IAEA2011].
+Body fat mass (FM, kg) was calculated by substituting LBM from body weight.
+The percentage of TBW, LBM, and FM were calculated based on body weight.
+Fat mass index (FMI) and Fat-free mass index (FFMI) were calculated using FMI = FM (kg) / height (m2) and FFMI = LBM (kg) / height (m2).
+With consent, venous blood samples were collected to measure leptin levels (ug/L).
+All measurements were conducted at the Institute of Infectious Diseases at Makerere University.
+
+## Immune response measures
+
+To assess immune activation, flow cytometry was used to measure CD4+ and CD8+ T cell counts in venous blood samples (7 ml of plasma).
+The percentages of CD4+ and CD8+ cells were calculated by dividing their counts by the total lymphocyte count.
+Levels of C-reactive protein (mg/dl), interferon-gamma (umol/ml), tumor necrosis factor-alpha (umol/ml), and interleukin-10 (ul/ml) were also measured as indicators of immune response.
+
+*Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
 @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
+We store those figures in the `assets` folder.*
 
@@
 
-## Data aquisition
-
-*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
-
-## Data import and cleaning
-
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
-
 ## Statistical analysis
 
-*Explain anything related to your statistical analyses.*
+*This part is under construction:* I will firstly compare body indices and immune response measures between males and females.
+As all variables of interest are continuous, I will use Wilcoxon rank sum test.
+To measure leptin’s effect in modulating immune activation and to link body energy storage with immune response strength, my initial plan is to use regression analysis to examine leptin’s mediator effect, but I’m seeking a more appropriate method.
+Of note, the sample size is relatively small, which may influence the significance of hypothesis tests.
 
@@
 
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
-
 {{< pagebreak >}}

### 2025-02-17 products/manuscript/Manuscript.qmd
Part 1 - 2
Organization

@@
 
-## General Background Information
-
 As a global threat, tuberculosis (TB) caused 10.8 million new cases in 2023 with Africa contributing 24% of these cases [@WHO2024].
@@
 The percentage of TBW, LBM, and FM were calculated based on body weight.
-Fat mass index (FMI) and Fat-free mass index (FFMI) were calculated using FMI = FM (kg) / height (m2) and FFMI = LBM (kg) / height (m2).
+Fat mass index (FMI) and Fat-free mass index (FFMI) were calculated using FMI = FM (kg) / height ($m^2$) and FFMI = LBM (kg) / height ($m^2$).
 With consent, venous blood samples were collected to measure leptin levels (ug/L).
@@
 Levels of C-reactive protein (mg/dl), interferon-gamma (umol/ml), tumor necrosis factor-alpha (umol/ml), and interleukin-10 (ul/ml) were also measured as indicators of immune response.
-
-*Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.*
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
 
@@
 
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
-```
-
 ## Basic statistical analysis
@@
 Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
-
-## Full analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
 

### 2025-02-21 code/processing-code/processingfile.qmd
Data cleaning

---
title: "Data cleaning demo"
author: "Guozheng Yang"
date: "2024-02-07"
output: html_document
---

# Processing script

This contains the same code and comments/information as `processingcode.R`.

But one can directly see some of the outputs.

# Setup

Load needed packages.

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
```

# Data loading

```{r}
# path to data
data_location <- here::here("data","raw-data","data0.xlsx")
rawdata <- readxl::read_excel(data_location)
```

# Check data

The codebook for this data is in the READMD file in the raw-data sub-folder.

An overview of the data structure.

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

# Cleaning

I will check the variables one by one.

*ID*: looks good

*Sex*: looks good

*Age*: looks good

*Height*: looks good

*Weight*: looks good

*BMI*: looks good

*TBW_L & TBW_pct*:

*TBW_L* only makes sense when it is lower than body weight. 
The *TBW_pct* should be within (0, 1).
Wrongly measured *TBW_L* will also result in weird read in other body indices.
Note that all individuals with *TBW_pct*>73% have negative *Fat_kg*.
Also, all individuals with *TBW_pct*<0 have *Fat_pct*>1. 
Here I keep all *TBW_pct*>73% or *TBW_pct*<0 and corresponding *TBW_L* as missing.
Other body composition indices will also be kept as NA.

```{r}
d1 <- rawdata %>%
  mutate(across(TBW_L:FFMI, ~ ifelse(TBW_pct > 0.73 | TBW_pct < 0, NA, .))) %>%
  mutate(across(ends_with("_pct"), ~ .*100)) # multiply percentages by 100
```

*Leptin_ng_ml*: looks good (one extremely high - 44+, but it's still reasonable)

*CRP_mg_dl*: looks good

*INF_Gamma_pg_ml*: looks good

*TNF_Alpha_pg_ml*: looks good

*IL_10_pg_mL*: looks good

*CD4_plus*: looks good

*CD4_plus_pct*: looks good (multiplied by 100 in last step)

*CD8_plus*: looks good

*CD8_plus_pct*: looks good (multiplied by 100 in last step)

All the other variables are fine. But I will round CD4+ counts and CD8+ counts to the nearest integer.

```{r}
d2 <- d1 %>%
  mutate(CD4_plus=round(CD4_plus),
         CD8_plus=round(CD8_plus))
```

The data cleaning process is completed. Let's take a final look at the cleaned data.

```{r}
dplyr::glimpse(d2)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

### 2025-02-21 products/manuscript/Manuscript.qmd
Data cleaning

@@
 # Methods
+
 ## Study design and participants
@@
 Participants who provided informed consent underwent total body water (TBW, L) estimation using deuterium dilution.
-Specifically, participants received a precisely weighed oral dose of deuterium oxide (~25-30g $D_2O$, 99.8% purity) after an overnight fast.
+Specifically, participants received a precisely weighed oral dose of deuterium oxide (\~25-30g $D_2O$, 99.8% purity) after an overnight fast.
 Participants will be asked to refrain from eating and to consume only the minimum amount of water needed to quench thirst during the 4-hour equilibration period.
@@
 Fat mass index (FMI) and Fat-free mass index (FFMI) were calculated using FMI = FM (kg) / height ($m^2$) and FFMI = LBM (kg) / height ($m^2$).
-With consent, venous blood samples were collected to measure leptin levels (ug/L).
+With consent, venous blood samples were collected to measure leptin levels (ng/ml).
 All measurements were conducted at the Institute of Infectious Diseases at Makerere University.
@@
 The percentages of CD4+ and CD8+ cells were calculated by dividing their counts by the total lymphocyte count.
-Levels of C-reactive protein (mg/dl), interferon-gamma (umol/ml), tumor necrosis factor-alpha (umol/ml), and interleukin-10 (ul/ml) were also measured as indicators of immune response.
+Levels of C-reactive protein (CRP, mg/dl), interferon-gamma (INF-gamma, pg/ml), tumor necrosis factor-alpha (TNF-alpha, pg/ml), and interleukin-10 (IL-10, pg/ml) were also measured as indicators of immune response.
 
@@
 *This part is under construction:* I will firstly compare body indices and immune response measures between males and females.
-As all variables of interest are continuous, I will use Wilcoxon rank sum test.
 To measure leptin’s effect in modulating immune activation and to link body energy storage with immune response strength, my initial plan is to use regression analysis to examine leptin’s mediator effect, but I’m seeking a more appropriate method.
@@
 
-## Exploratory/Descriptive analysis
+## Difference in body composition index by sex
 
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
+The 30 male and 30 female participants had a similar age distribution (@tbl-body).
+Males were taller (median = 168 cm, IQR: 158 cm-175 cm) and heavier (median = 55.2 kg, IQR: 50.0 kg-60.0 kg) than females, but BMI was similar between sexes (median: 19.48 vs.
+19.20).
+Males had a higher body water proportion (median = 63.67%, IQR: 59.92%-68.09%) and lean body mass composition (median = 86.98%, IQR: 81.86%-93.01%), while females had a higher body fat composition (median = 21.37%, IQR: 14.74%-27.09%).
+After adjusting for height, males had a lower FMI (median: 2.46 vs.
+4.41) but a higher FFMI (median: 16.80 vs.
+15.19) than females.
+Leptin levels were higher in females than in males (median: 1.16 vs.
+1.56).
 
-@tbl-summarytable shows a summary of the data.
+```{r}
+#| label: tbl-body
+#| tab-cap: "Body characteristics and composition by sex."
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","table1.png"))
+```
 
-## Basic statistical analysis
+## Difference in immune response by sex
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
+CRP, INF-gamma, TNF-alpha, and IL-10 levels showed no clear differences between males and females (@fig-boxplot).
+However, females had more compact distributions for CRP, TNF-alpha, and IL-10 levels.
+Both sexes had INF-gamma levels well above zero.
+Lymphocyte subsets varied between sexes.
+CD4+ levels were lower in males but showed less variation.
+The difference was not evident regarding CD8+ level.
+Among males, the median CD4+ count was around 300/uL, with a median percentage of about 20%, while females had a median count above 400/uL.
+Both sexes had CD4+ and CD8+ counts as low as nearly zero.
+The maximum CD4+ count reached 800/uL, whereas the maximum CD8+ count was over 400/uL in males and nearly 700/uL in females.
+
+```{r}
+#| label: fig-boxplot
+#| fig-cap: "Comparison of immune response by sex. The boxplots denote immune response measures among male (blue) and female (orange) participants. "
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","figure1.png"))
+```
+
+## Association between body composition and immune response
+
+Leptin levels were strongly associated with body fat among females, with a positive correlation (Pearson correlation coefficient = 0.6, @fig-sct).
+In contrast, leptin levels showed a negative correlation with lean body mass in females (Pearson correlation coefficient = -0.4).
+Males had consistently lower leptin levels than females, with no evident correlation with body fat or lean body mass.
+No clear association was observed between leptin levels and lymphocyte subsets.
+The correlation between leptin levels and the percentage of CD4+ cells was weak (Pearson correlation coefficient = 0.09).
+Similarly, the correlation with the percentage of CD8+ cells was slightly negative but not significant (Pearson correlation coefficient = -0.13).
+
+```{r}
+#| label: fig-sct
+#| fig-cap: "Association between log-transformed leptin level and body fat, lean body mass, percentage of CD4+ cells, and percentage of CD8+ cells.
+Male and female participants are denoted by blue and pink points, respectively.
+The size of point denotes body weight of each participant.
+R marks the Pearson correlation coefficients.
+"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","figure2.png"))
+```
+
+## Mediation effect of leptin in immune response activation
 
@@
 # Discussion
-
-## Summary and Interpretation
-
-*Summarize what you did, what you found and what it means.*
-
-## Strengths and Limitations
-
-*Discuss what you perceive as strengths and limitations of your analysis.*
 

### 2025-02-21 code/analysis-code/statistical-analysis.R

@@ -5,52 +5,294 @@
 #and saves the results to the results folder
 
 #load needed packages. make sure they are installed.
-library(ggplot2) #for plotting
-library(broom) #for cleaning up output from lm()
+library(tidyverse) # for data analysis
+library(gt) # for making tables
+library(gtsummary) # for summary table
+library(ggplot2) # for plotting
+library(ggpubr) # for combining sub-figures
 library(here) #for data loading/saving
 
 #path to data
 #note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","processeddata.rds")
 
-#load data. 
-mydata <- readRDS(data_location)
+# Load data. 
+data <- readRDS(data_location)
+head(data)
+
+# Define the order of sex
+data$Sex <- factor(data$Sex, levels=c("M", "F"))
+
+################################################################################
+# Comparing body indices  between Male vs Female
+################################################################################
+
+# Make a summary table 
+summary_tab <- data %>% select(-c(ID, CRP_mg_dl, INF_Gamma_pg_ml, TNF_Alpha_pg_ml, IL_10_pg_mL,
+                                  CD4_plus, CD4_plus_pct, CD8_plus, CD8_plus_pct)) %>%
+  tbl_summary(by=Sex, type=list(where(is.numeric) ~ "continuous"),
+              statistic=list(all_continuous() ~ "{median} ({p25}, {p75})"),
+              digits=all_continuous() ~ 2, missing="no",
+              label=list(Age ~ "Age, years",
+                         Height ~ "Height, cm",
+                         Weight ~ "Weight, kg",
+                         BMI ~ "Body mass index",
+                         TBW_L ~ "Body water, L",
+                         TBW_pct ~ "Percentage of body water, %",
+                         Fat_kg ~ "Body fat, kg",
+                         Fat_pct ~ "Percentage of body fat, %",
+                         FMI ~ "Fat mass index",
+                         LBM_kg ~ "Lean body mass, kg",
[…]

### 2025-02-21 code/processing-code/processingcode.R

@@ -1,12 +1,9 @@
 ###############################
 # processing script
 #
-#this script loads the raw data, processes and cleans it 
-#and saves it as Rds file in the processed-data folder
+# this script loads the raw data, processes and cleans it 
+# and saves it as Rds file in the processed-data folder
 #
-# Note the ## ---- name ---- notation
-# This is done so one can pull in the chunks of code into the Quarto document
-# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
 ###############################
 
 
@@ -22,7 +19,7 @@ library(here) #to set paths
 ## ---- loaddata --------
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw-data","exampledata.xlsx")
+data_location <- here::here("data","raw-data","data0.xlsx")
 #load data. 
 #note that for functions that come from specific packages (instead of base R)
 # I often specify both package and function like so
@@ -30,109 +27,74 @@ data_location <- here::here("data","raw-data","exampledata.xlsx")
 #specifying the package makes it clearer where the function "lives",
 #but it adds typing. You can do it either way.
 rawdata <- readxl::read_excel(data_location)
-# We might also want to load the codebook to look at it
-codebook <- readxl::read_excel(data_location, sheet ="Codebook")
 
+# The codebook for this data is in the READMD file in the raw-data sub-folder.
 
 ## ---- exploredata --------
-#take a look at the data
+# Take a look at the data
 dplyr::glimpse(rawdata)
-#another way to look at the data
+# Look at the distribution
 summary(rawdata)
-#yet another way to get an idea of the data
+# Take a look at the first several rows
 head(rawdata)
-#this is a nice way to look at data
+# Another way to look at the distribution
 skimr::skim(rawdata)
-#look in the Codebook for a variable explanation
-print(codebook)
 
 
-## ---- cleandata1 --------
-# Inspecting the data, we find some problems that need addressing:
-# First, there is an entry for height which says "sixty" instead of a number. 
-# Does that mean it should be a numeric 60? It somehow doesn't make
-# sense since the weight is 60kg, which can't happen for a 60cm person (a baby)
-# Since we don't know how to fix this, we might decide to remove the person.
[…]

### 2025-03-14 products/manuscript/Manuscript.qmd
Main analysis - 1

@@
 
-*This part is under construction:* I will firstly compare body indices and immune response measures between males and females.
-To measure leptin’s effect in modulating immune activation and to link body energy storage with immune response strength, my initial plan is to use regression analysis to examine leptin’s mediator effect, but I’m seeking a more appropriate method.
-Of note, the sample size is relatively small, which may influence the significance of hypothesis tests.
-
 {{< pagebreak >}}
@@
 #| label: tbl-body
-#| tab-cap: "Body characteristics and composition by sex."
+#| tbl-cap: "Body characteristics and composition by sex."
 #| echo: FALSE
@@
 
+(@tbl-coef)
+
+```{r}
+#| label: tbl-coef
+#| tbl-cap: "Coefficient estimation (95% CI) from linear regression and Poisson regression."
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","table2.png"))
+```
+
+(@tbl-ma)
+
+```{r}
+#| label: tbl-ma
+#| tbl-cap: "Mediation analysis with estimated effects and 95% CI."
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","table3.png"))
+```
+
 {{< pagebreak >}}

### 2025-03-14 code/analysis-code/statistical-analysis.R

@@ -4,13 +4,16 @@
 #this script loads the processed, cleaned data, does a simple analysis
 #and saves the results to the results folder
 
+rm(list = ls())
+
 #load needed packages. make sure they are installed.
 library(tidyverse) # for data analysis
 library(gt) # for making tables
 library(gtsummary) # for summary table
 library(ggplot2) # for plotting
 library(ggpubr) # for combining sub-figures
-library(here) #for data loading/saving
+library(here) # for data loading/saving
+library(boot) # for bootstrapping: estimate CI
 
 #path to data
 #note the use of the here() package and not absolute paths
@@ -28,8 +31,8 @@ data$Sex <- factor(data$Sex, levels=c("M", "F"))
 ################################################################################
 
 # Make a summary table 
-summary_tab <- data %>% select(-c(ID, CRP_mg_dl, INF_Gamma_pg_ml, TNF_Alpha_pg_ml, IL_10_pg_mL,
-                                  CD4_plus, CD4_plus_pct, CD8_plus, CD8_plus_pct)) %>%
+summary_tab <- data %>% dplyr::select(-c(ID, CRP_mg_dl, INF_Gamma_pg_ml, TNF_Alpha_pg_ml, IL_10_pg_mL,
+                                         CD4_plus, CD4_plus_pct, CD8_plus, CD8_plus_pct)) %>%
   tbl_summary(by=Sex, type=list(where(is.numeric) ~ "continuous"),
               statistic=list(all_continuous() ~ "{median} ({p25}, {p75})"),
               digits=all_continuous() ~ 2, missing="no",
@@ -61,7 +64,7 @@ summary_tab <- data %>% select(-c(ID, CRP_mg_dl, INF_Gamma_pg_ml, TNF_Alpha_pg_m
               table.border.top.color="transparent",
               table.border.bottom.color="black",
               table.border.bottom.width=3,
-              table.width=pct(80)) %>%
+              table.width=pct(60)) %>%
   tab_style(style=cell_borders(sides="bottom", color="black", weight=px(2)), locations=cells_column_labels()) %>%
[…]

### 2025-03-15 products/manuscript/Manuscript.qmd
Main analysis - 2

@@
 
-{{< pagebreak >}}
-
 # Methods
@@
 
-{{< pagebreak >}}
+In the descriptive analysis, we compared the distribution of demographic characteristics (e.g., age, height, weight) and body composition indices (e.g., body fat, lean body mass, leptin level) between male and female participants using the median and interquartile range (IQR).
+Differences in immune responses were visualized with boxplots showing the distributions of available measures, including CD4+ cell counts, CRP levels, and INF-gamma levels.
+The association between log-transformed leptin levels and body fat, lean body mass, percentage of CD4+ cells, and percentage of CD8+ cells was assessed using scatterplots with Pearson correlation coefficients.
+
+We examined the predictability of fat mass index to leptin level using a simple linear regression.
+A Poisson regression model was applied to CD4+ cell counts using a log link, with fat mass index and leptin level as predictors.
+All leptin levels in regression models used the logarithmic transformation.
+We estimated standardized regression coefficients and calculated the total, direct, and indirect effects of fat mass index on CD4+ cell counts [@Sobel1982].
+The mediation effect of leptin was quantified as the indirect effect divided by the total effect, expressed as the percentage of the total effect mediated by leptin [@Baron1986].
+Confidence intervals (CI) at the 95% confidence level for all estimated effects were obtained using the bootstrap method with 500 replicates.
+The mediation analysis was first conducted for all participants.
+We then performed a subgroup analysis by applying the mediation analysis separately to male and female participants.
 
@@
 
-(@tbl-coef)
+When fitting linear regression for all participants, leptin level was positively associated with FMI ($\beta$=0.100, 95% CI: 0.064-0.135, @tbl-coef).
+A similar positive association was observed among female participants ($\beta$=0.113, 95% CI: 0.062-0.165), but not among males ($\beta$=0.000, 95% CI: -0.003-0.002).
+When predicting CD4+ cell counts using FMI and leptin level for all participants, both predictors showed a positive association ($\beta$=0.008, 95% CI: 0.000-0.017 for FMI; $\beta$=0.364, 95% CI: 0.314-0.413 for leptin level).
+Among males, both FMI ($\beta$=0.025, 95% CI: 0.011-0.040) and leptin level ($\beta$=6.995, 95% CI: 4.489-9.500) were positively associated with CD4+ cell counts.
+However, among females, CD4+ cell counts were positively associated only with leptin level ($\beta$=0.322, 95% CI: 0.262-0.381).
 
@@
 
-(@tbl-ma)
+Among all participants, FMI had a notable effect on promoting CD4+ cell counts (total effect=0.25, 95% CI: -0.15-0.57, @tbl-ma), with a large proportion (indirect effect=0.20, 95% CI: -0.14-0.55) mediated by leptin level.
+The mediation effect of leptin accounted for 82% of the total effect of FMI.
+A similar pattern was observed among female participants, where the effect of FMI (total effect=0.24, 95% CI: -0.35-0.67) was almost entirely mediated by leptin (indirect effect=0.24, 95% CI: -0.30-0.80).
+Among male participants, CD4+ cell counts were primarily influenced by the direct effect of FMI (total effect=0.09, 95% CI: -0.55-0.76; direct effect=0.10, 95% CI: -0.57-0.89).
 
@@
 
-{{< pagebreak >}}
-
 # Discussion

### 2025-03-15 products/manuscript/Manuscript.qmd
Abstract

@@
 
-# Abstract
+# Abstract {.unnumbered}
 
-## Background:
+## Background: {.unnumbered}
 
-## Methods:
+In tuberculosis infection, higher body fat storage in females may better support immune activation and response compared to males.
+Leptin may mediate this process, bridging metabolism and the immune system.
 
-## Results:
+## Methods: {.unnumbered}
 
-## Conclusions:
+We conducted a cross-sectional study in Kampala, Uganda, recruiting 60 participants with the first episode of tuberculosis.
+Demographic characteristics, body composition indices, and immune response measures were compared between 30 male and 30 female participants.
+A structural equation-based mediation analysis examined the association between CD4+ cell levels and body fat, as well as leptin’s mediation effect.
+The analysis was performed for all participants and separately by sex.
 
-# Keywords:
+## Results: {.unnumbered}
 
-*Write a summary of your project.*
+Compared to male participants, females had a higher fat mass index (median: 2.46 vs.
+4.41), higher leptin levels (median: 1.16 ng/ml vs.
+1.56 ng/ml), and higher CD4+ cell levels.
+Leptin levels were positively associated with body fat storage, with a stronger effect observed in females.
+Among all participants, 82% of the effect of body fat storage on immune activation was mediated by leptin, with a more pronounced mediation effect in females than in males.
+
+## Conclusions: {.unnumbered}
+
+Enhancing body fat storage may support immune activation in tuberculosis infection. Leptin could serve as a therapeutic target for sex-specific immune modulation in tuberculosis.
+
+# Keywords: {.unnumbered}
+
+**Tuberculosis, leptin, body fat, immune response**
 
@@
 15.19) than females.
-Leptin levels were higher in females than in males (median: 1.16 vs.
-1.56).
+Leptin levels were higher in females than in males (median: 1.16 ng/ml vs.
+1.56 ng/ml).
 
@@
 
-## Conclusions
-
 {{< pagebreak >}}
 
-# References
+# References {.unnumbered}

### 2025-03-15 products/manuscript/Manuscript.qmd
Discussion

@@
 
+This study examined the role of leptin in metabolism and immune response regulation.
+The sample included 30 males and 30 females with the first TB episode from Uganda.
+Our findings revealed that females had higher body fat storage, higher leptin levels, and stronger immune responses.
+Body fat was positively associated with leptin levels, and leptin mediated the effect of body fat on immune activation.
+
+Our findings align with previous studies showing that females tend to have a stronger immune response than males [@Klein2016].
+This difference in immunity may influence susceptibility and the speed of infection clearance between sexes [@Takahashi2021].
+This study focused on immune activation and identified a sex-based difference in CD4+ cell levels, consistent with findings from a previous study [@Bothamley2021].
+Recent research also indicates that changes in body fat in males can significantly reduce CD8+ cell levels compared to females [@Dhanyalayam2024].
+Our study, along with previous research, provides evidence of sex-specific differences in the initiation and dynamic response of the immune system to TB infection.
+Our study also identified leptin as a bridge between metabolism and the immune system.
+This suggests its potential as a therapeutic target for regulating immune responses based on specific metabolic conditions [@Naylor2016].
+Animal-based preclinical studies have demonstrated its practical significance [@Fantuzzi2000], but further research is needed to assess its clinical applicability in human populations.
+
+This study has several limitations.
+First, the small sample size reduced the statistical power of the analysis.
+Sample loss due to the instability of the deuterium dilution approach in measuring body water further limited the detection of significant associations.
+Second, non-biological factors such as professional exposure, alcohol use, and smoking status [@Wigger2022] were not controlled, which could have influenced immune responses.
+Given their higher prevalence among males [@Soh2017], these factors may have confounded the findings.
+Despite these limitations, this study provides valuable insights that complement previous research.
+Future studies should explore the dynamic regulatory role of leptin at different stages of immune responses.
+
 {{< pagebreak >}}

### 2025-04-13 products/manuscript/Manuscript.qmd
ML - part 4

@@
 The analysis was performed for all participants and separately by sex.
+To incorporate material from this course, we further evaluated the performance of random forest and regression tree models in predicting leptin levels using body composition indices.
+One-fifth of the data set was reserved as a test set for model evaluation, while hyperparameter tuning was conducted on the training set using five-fold cross-validation.
 
@@
 Among all participants, 82% of the effect of body fat storage on immune activation was mediated by leptin, with a more pronounced mediation effect in females than in males.
+The random forest model is better than the regression tree in predicting leptin levels using body composition indices, with a lower root mean squared error (0.40 vs 0.41) and a higher R-square (0.53 vs.
+0.23) on the test set.
 
@@
 Specifically, participants received a precisely weighed oral dose of deuterium oxide (\~25-30g $D_2O$, 99.8% purity) after an overnight fast.
-Participants will be asked to refrain from eating and to consume only the minimum amount of water needed to quench thirst during the 4-hour equilibration period.
+Participants were asked to refrain from eating and to consume only the minimum amount of water needed to quench thirst during the 4-hour equilibration period.
 Saliva samples collected before and after a 4-hour equilibration period were analyzed using Fourier transform infrared spectrophotometry to determine deuterium concentration, and TBW was estimated by extrapolating deuterium dilution space, adjusted by a factor of 1.041 [@IAEA2011].
@@
 
+We further compared the performance of random forest and regression tree models in predicting log-transformed leptin levels using body composition indices.
+The predictors included gender, age, BMI, percentage of lean body mass, and percentage of fat mass.
+One-fifth of the dataset was set aside as the test set, with the remaining data used for training.
+Five-fold cross-validation was applied to the training set for hyperparameter tuning.
+For the random forest model, a grid search was conducted to optimize the number of randomly sampled predictors and the minimum number of observations required to split a node, targeting the lowest root mean squared error (RMSE).
+For the regression tree model, cost complexity and tree depth were tuned to minimize RMSE as well.
+Model performance was evaluated on the test set using RMSE and R-squared.
+
 # Results
@@
 The 30 male and 30 female participants had a similar age distribution (@tbl-body).
-Males were taller (median = 168 cm, IQR: 158 cm-175 cm) and heavier (median = 55.2 kg, IQR: 50.0 kg-60.0 kg) than females, but BMI was similar between sexes (median: 19.48 vs.
+Males were taller (median = 168.00 cm, IQR: 158.00 cm-175.00 cm) and heavier (median = 55.20 kg, IQR: 50.00 kg-60.00 kg) than females, but BMI was similar between sexes (median: 19.48 vs.
 19.20).
@@
 
+## Prediction leptin level using body compositions
+
+The tuning results indicate that the random forest model achieved optimal performance with five randomly sampled predictors and a minimum split size of approximately 15 (@fig-rf).
+This led to an RMSE of 0.40 and an R-square of 0.53 on the test set.
+In contrast, the regression tree model demonstrated minimal variation in RMSE across different values of tree depth and cost complexity (@fig-rt), with the optimal parameter set resulting in an RMSE of 0.41 and an R-square of 0.23 in the test set.
+This lack of sensitivity to parameter changes suggests that the regression tree may have reached its performance ceiling due to the limited sample size and the model's inherent simplicity.
+Unlike random forest, which benefits from averaging multiple trees and capturing more complex patterns, the single-tree structure of the regression tree model may not sufficiently capture the underlying variability in the data.
+
+```{r}
+#| label: fig-rf
+#| fig-cap: "Parameter tuning results of random forest. "
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","figure3.png"))
+```
+
+```{r}
+#| label: fig-rt
+#| fig-cap: "Parameter tuning results of regression tree. "
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","figure4.png"))
+```
+
+The plots comparing predicted values and observed values further illustrate the performance differences between the two models (@fig-rfrt).
+Predicted values from the random forest model are more clustered, generally following the diagonal line.
+This indicates a better agreement between predicted and observed leptin levels.
+In contrast, the regression tree model shows clear signs of underfitting, with predicted values appearing in discrete horizontal bands and failing to capture the continuous nature of leptin levels.
+This is one of the characteristics of single-tree models, which produces piece-wise constant predictions and lacks the flexibility to model associations.
+Overall, the random forest model provides more accurate and nuanced predictions, benefiting from its more flexible structure and ability to reduce variance.
+
+```{r}
+#| label: fig-rfrt
+#| fig-cap: "Comparison of random forest and regression tree: predicted values vs. observed values. "
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","figure5.png"))
+```
+
 # Discussion
@@
 Given their higher prevalence among males [@Soh2017], these factors may have confounded the findings.
+the machine learning component of this study serves primarily as a methodological exercise due to the limited sample size.
+A substantially larger data set would be required to enable a more robust and convincing comparison between the random forest and regression tree models.
+However, recruiting a sufficiently large sample of patients experiencing their first episode of TB in high-prevalence regions remains a significant challenge.
 Despite these limitations, this study provides valuable insights that complement previous research.

### 2025-04-13 code/analysis-code/ML-analysis.R

@@ -1,0 +1,177 @@
+###############################
+# Machine Learning Script
+#
+#this script applies ML methods to analyze the data set.
+#Performances of regression tree and random forest are compared. 
+
+rm(list = ls())
+
+#load needed packages. make sure they are installed.
+library(tidyverse) # for data analysis
+library(gt) # for making tables
+library(gtsummary) # for summary table
+library(ggplot2) # for plotting
+library(ggpubr) # for combining sub-figures
+library(here) # for data loading/saving
+library(tidymodels) # for ML models
+
+#path to data
+#note the use of the here() package and not absolute paths
+data_location <- here::here("data","processed-data","processeddata.rds")
+
+# Load data. 
+data <- readRDS(data_location)
+head(data)
+
+# Define the sex variable: 1-Female
+data$Female <- ifelse(data$Sex=="F", 1, 0)
+# Convert scale of leptin
+data$log_leptin <- log10(data$Leptin_ng_ml)
+
+# Select predictors for ML models
+data <- data %>% select(log_leptin, Female, Age, BMI, LBM_pct, Fat_pct)
+
+# Set a seed
+rngseed <- 1234
+
+# 5-fold CV
+data_split <- initial_split(data, prop=.8)
+train_data <- training(data_split)
+test_data  <- testing(data_split)
+folds <- vfold_cv(train_data, v=5)
+
+################################################################################
+# Random Forest model
+################################################################################
+
+# Set seed
+set.seed(rngseed)
+
+# Random forest model
+rcp <- recipe(log_leptin ~ Female + Age + BMI + LBM_pct + Fat_pct, data=train_data) %>%
+  step_normalize(all_numeric_predictors())
+
+# Add flexibility: tune the two parameters
+rf_spec <- rand_forest(mtry=tune(), min_n=tune(), trees=10) %>%
+  set_engine("ranger") %>%
+  set_mode("regression")
+rf_wf <- workflow() %>% add_model(rf_spec) %>% add_recipe(rcp)
+
+# Tune the two parameters based on RMSE
+rf_tune <- tune_grid(rf_wf, resamples=folds, grid=10, metrics=metric_set(rmse))
+
+# Make a plot of tuning results
+rf_tune_df <- as.data.frame(rf_tune$.metrics)
+rf_tune_plot <- ggplot(rf_tune_df, aes(x=mtry, y=min_n, fill=.estimate))+
+  geom_tile()+
[…]

### 2025-04-14 products/manuscript/Manuscript.qmd
ML - part 4 - simulation

@@
 
+Last, we simulated a data set with log-transformed leptin level, sex, age, and BMI.
+This simulated data set was then used to examine the performance of random forest model in predicting leptin level using the other variables as predictors.
+The sex variable was generated using a Bernoulli distribution with a probability of 0.6 of being females.
+Age was assumed to follow a normal distribution with a mean of 35 and standard deviation of 3.
+BMI was also assumed a normal distribution with a mean of 25 and standard deviation of 1.5.
+We generated log-transformed leptin levels assuming the relationship between it and the other predictors as: $log-leptin=8*Sex-0.1*Age^2+6*Age+exp(0.05*BMI)$.
+Besides the association structure as described above, we also added an error term following a standard normal distribution.
+The model evaluation process was the same as conducted on the real-world data set.
+
 # Results
@@
 
+# Simulation study using random forest
+
+The performance of the random forest model on the simulated data set appears to be highly accurate and optimized (@fig-rfsim, @fig-rfpredsim).
+The parameter tuning results demonstrate that the lowest RMSE was achieved when the number of randomly sampled predictors was set to 3 and the minimum number of split points was 37.
+In contrast, the model exhibited higher RMSEs as the number of randomly sampled predictors decreased, suggesting that incorporating a larger subset of predictors improves predictive performance.
+Additionally, the scatterplot comparing predicted and observed values shows that most points lie very close to the diagonal line, indicating strong agreement between predictions and actual outcomes.
+These findings collectively suggest that the random forest model effectively captured the complex, non-linear relationships embedded in the simulated data, resulting in robust and reliable predictions.
+
+```{r}
+#| label: fig-rfsim
+#| fig-cap: "Parameter tuning results of random forest model using simulated data. "
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","figure6.png"))
+```
+
+```{r}
+#| label: fig-rfpredsim
+#| fig-cap: "Fitting results of random forest on simulated data: predicted values vs. observed values. "
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","figure7.png"))
+```
+
 # Discussion
@@
 Animal-based preclinical studies have demonstrated its practical significance [@Fantuzzi2000], but further research is needed to assess its clinical applicability in human populations.
+The extended two sections demonstrate the better performance of the random forest model in predicting leptin levels using body composition indices.
+Its strong performance is further demonstrated when non-linear relationships exist.
 

### 2025-04-14 code/analysis-code/simulation-analysis.R

@@ -1,0 +1,114 @@
+###############################
+# Simulation Script
+#
+#this script applies the RF model to simulated data set.
+
+rm(list = ls())
+
+#load needed packages. make sure they are installed.
+library(tidyverse) # for data analysis
+library(ggplot2) # for plotting
+library(ggpubr) # for combining sub-figures
+library(here) # for data loading/saving
+library(tidymodels) # for ML models
+
+# Simulate data set: N=1000
+N <- 1000
+# log_leptin ~ 8*Female + -0.1*Age^2 + 6*Age + exp(0.05*BMI) + error
+# Female ~ Bernoulli(0.6)
+# Age ~ N(35, 3)
+# BMI ~ N(25, 1.5)
+# error ~ N(0,1)
+
+# Set a seed
+rngseed <- 1234
+set.seed(rngseed)
+
+# Simulate a data set
+data <- data.frame(Female=rbinom(N, size=1, prob=.6),
+                   Age=rnorm(N, 35, 3),
+                   BMI=rnorm(N, 25, 1.5),
+                   error=rnorm(N, 0, 1))
+data$log_leptin <- 8*data$Female + -0.1*data$Age^2 + 6*data$Age + exp(0.05*data$BMI) + data$error
+
+# Select predictors for RF
+data <- data %>% select(log_leptin, Female, Age, BMI)
+
+# Set a seed
+set.seed(rngseed)
+# 5-fold CV
+data_split <- initial_split(data, prop=.8)
+train_data <- training(data_split)
+test_data  <- testing(data_split)
+folds <- vfold_cv(train_data, v=5)
+
+################################################################################
+# Random Forest model
+################################################################################
+
+# Set seed
+set.seed(rngseed)
+
+# Random forest model
+rcp <- recipe(log_leptin ~ Female + Age + BMI, data=train_data) %>%
+  step_normalize(all_numeric_predictors())
+
+# Add flexibility: tune the two parameters
+rf_spec <- rand_forest(mtry=tune(), min_n=tune(), trees=50) %>%
+  set_engine("ranger") %>%
+  set_mode("regression")
+rf_wf <- workflow() %>% add_model(rf_spec) %>% add_recipe(rcp)
+
+# Tune the two parameters based on RMSE
+rf_tune <- tune_grid(rf_wf, resamples=folds, grid=10, metrics=metric_set(rmse))
+
+# Make a plot of tuning results
[…]

### 2025-04-25 Review - Gokhale/project-review-yang.qmd
Review - Gokhale

---
title: Project Review  
author: Pooja Gokhale
date: April-23-2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Sex-Based Differences in Tuberculosis Immune Response: The Role of Leptin

Name of project author(s): Guozheng Yang

Name of project reviewer: Pooja Gokhale

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/renders into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
### Feedback and Comments

The background is well structure, and provides appropriate information. Africa has specifically been mentioned since the data is taken from Uganda, however, the introduction could benefit from further elaboration. Perhaps mention in the aim that the data was taken from Africa, tying it back together with the first paragraph.

'Since women generally have higher body fat storage, their immune systems may receive sufficient energy to support a stronger response to infections compared to men. In diseases like TB that often cause body wasting (7), differences in energy reserves may contribute to variations in immune response between men and women.' - It would be helpful to provide a reference for this statement.
If this is a hypothesis, and no reference is available, this should be stated as such.

The further paragraphs state that the role of leptin does not have enough evidence. However, there is no evidence provided regarding immune responses in general in women.

### Summary assessment 
* strong contextualization and motivation

## Question description

### Feedback and Comments

The question has been clearly stated.

### Summary assessment
* question/hypotheses fully clear

## Data description
### Feedback and Comments
The data is well described, as well as the source.
The different measures, and how they are obtained has been clearly described.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis
### Feedback and Comments
The data cleaning process has not been explained at all in the manuscript. The processing code file shows minor data cleaning.
Since the data was relatively clean already, minor data cleaning was needed.
It would be helpful to explain this part either in the supplementary material or in the main manuscript.
The exploratory analysis is explained well in the manuscript.
It would be helpful to include this analysis as a separate eda-code file, rather than having all the analysis in the statistical analysis file.

### Summary assessment
* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

### Feedback and Comments
The analysis methods are appropriate.
However, a multivariable linear regression model could be considered as well as a simple linear regression model.

### Summary assessment
* strong and reasonable analysis

## Presentation
### Feedback and Comments
In general, the figures and tables are publication-quality.
Table 2 is slightly confusing - it might be better to have separate tables for linear and poisson regression.
Additionally, consider a more thorough explanation of the 95% CI (for example - what is significant, add p-values).

### Summary assessment
* results are presented ok, with room for improvement

## Discussion/Conclusions

### Feedback and Comments
The study findings are explained properly.
The limitations have also been discussed appropriately.

### Summary assessment
* strong, complete and clear discussion

## Further comments
Overall, a very strong project. Some minor changes with writing and presentation of results would help improve clarity.

# Overall project content evaluation
## Structure
### Feedback and Comments

The project is well structured.
Files are correctly labeled - with reasonable names.
A separate file for exploratory data analysis would be helpful.
All readme files have been updated.

### Summary assessment
* well structured

## Documentation
### Feedback and Comments
Project is generally well documented. However, there are not enough comments to understand the code completely.

 ### Summary assessment
* decently documented with some gaps

## Reproducibility

### Feedback and Comments
The results are full reproducible. The readme files explain how to run the analysis. No manual edits needed.

### Summary assessment
* fully reproducible without issues

## Thoroughness

### Feedback and Comments
Alternative models were not considered or discussed. The hypothesis was answered, assessing the association between fat, leptin and immune response. However, a multivariable linear regression would have provided more valuable information. For example, would the results change if body water was also considered as a covariate?

### Summary assessment
* decent level of thoroughness

### Further comments
Consider a multivariable linear regression model, using all available relevant predictors. 
For example - sex, age, fmi, body water as predictors for leptin.

### 2025-04-28 Review - Grismer/yang-review-byhopeg.qmd
Review - Grismer

---
title: Project Review Template 
author: Hope Grismer
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Sex-Based Differences in Tuberculosis Immune Response: The Role of Leptin

Name of project author(s): G. Yang

Name of project reviewer: Hope Grismer

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/renders into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

# Specific project content evaluation

## Background, Context and Motivation

How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The background is strong and clear, placing the project into the broader context of tuberculosis, sex differences in immune responses, and the metabolic role of leptin.\
The motivation for the study is well stated, though it would be even stronger if the introduction included more direct references or citations supporting the specific immune differences between sexes (besides leptin pathways).

Suggestion: If no direct references exist, explicitly framing this as a hypothesis would improve clarity.

### Summary assessment (PICK ONE, DELETE THE OTHERS)

-   strong contextualization and motivation

## Question description

How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The research questions and hypotheses are clearly articulated:

-   Women may have stronger immune responses during TB infection due to higher body fat/leptin.

-   Leptin mediates the link between body composition and immune activation.

They are closely aligned with the data being collected and analyzed.

### Summary assessment

-   question/hypotheses fully clear

## Data description

How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The data sources and measurements are very well explained. The manuscript is transparent about how all variables were generated, and methods are described sufficiently to replicate measurements.

### Summary assessment

-   source and overall structure of data well explained

## Data wrangling and exploratory analysis

How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

The exploratory analysis in the manuscript is strong, with thoughtful visualizations and descriptive statistics.

However, data wrangling (cleaning) is barely described in the manuscript itself.\
From the processing script, it is clear that minimal cleaning was necessary — but that context should still be briefly mentioned in the manuscript or supplemental materials.\
Also, it would be better practice to have the exploratory plots in a separate EDA script (instead of combining all statistical analysis into one giant `.R` file).

### Summary assessment

-   some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

Minor suggestion: Including a multivariable linear model adjusting for possible confounders (e.g., age, sex, body composition) could have strengthened the findings, especially for leptin prediction.

But overall, analysis choices are sound and competently implemented.

### Summary assessment

-   strong and reasonable analysis

## Presentation

How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

The figures and tables are clear and easy to interpret.

However, a few suggestions:

-   Table 2 is confusing because it combines linear and Poisson regression coefficients.\
    ➔ Better to split into two separate tables.

-   P-values should be included alongside 95% confidence intervals for clarity.

-   Explain significance thresholds (e.g., p \< 0.05) explicitly in captions or footnotes.Summary assessment

<!-- -->

-   results are very well presented (tiny room for improvement)

### Summary assessment

-   results are presented ok, with room for improvement

## Discussion/Conclusions

Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The discussion is strong, thoughtful, and places findings in a wider biological and public health context.

The link between findings and broader TB immunology research is made very well.

Minor suggestion: Future research directions could be elaborated slightly more (e.g., larger cohorts, inclusion of HIV+ individuals).

### Summary assessment

-   strong, complete and clear discussion

## Further comments

Slightly better description of cleaning/EDA would make it even stronger.

# Overall project content evaluation

Evaluate overall features of the project by filling in the sections below.

## Structure

Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

-   Three main Quarto files corresponding to cleaning, EDA, and analysis

-   No junk files

-   Clear naming conventions

### Summary assessment

-   Well structured

## Documentation

How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

The manuscript itself is very well documented.\
The code is mostly documented, but:

-   Some sections lack comments explaining the purpose of each chunk

-   More small comments explaining model fitting choices would improve readability.

### Summary assessment

-   decently documented with some gaps

## Reproducibility

Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

-   All code files are provided and it is reproducible

### Summary assessment

-   fully reproducible without issues

## Thoroughness

How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

The project is thorough with:

-   Mediation models

-   Machine learning prediction models

-   A simulation study to validate methods

However, alternative regression models (e.g., multivariable modeling adjusting for covariates) could have been explored a bit more.

### Summary assessment

-   strong level of thorougness (with minor suggestions)

## Further comments

Thank you for an amazing project! I really was inspired in getting to look yours over. It was super well-rounded project– very close to publication-level. I especially liked the depth and strength of writing in your manuscript.

### 2025-05-02 products/manuscript/Manuscript.qmd
Final Change

@@
 
+## Data cleaning
+
+Due to measurement instability, a small number of participants had total body water (TBW) values recorded as less than 0 or greater than 0.70, which are biologically implausible.
+As TBW serves as a foundational parameter for calculating other body composition indices, participants with TBW values outside the plausible range (<0 or >0.73) were excluded from downstream analyses to ensure data quality and the validity of derived measures.
+In addition, percentage-based indices such as body water and body fat were originally recorded in decimal format and were converted to percentage format (e.g., 0.45 to 45%) to enhance interpretability.
+CD4+ and CD8+ cell counts were rounded to the nearest integer to reflect the typical reporting standard in clinical and immunological research.
+
 ## Statistical analysis
@@
 #| label: tbl-coef
-#| tbl-cap: "Coefficient estimation (95% CI) from linear regression and Poisson regression."
+#| tbl-cap: "Coefficient estimation (95% CI) from linear regression and Poisson regression.
+Fat mass index is only included in the linear regression model, while both predictors are included in the Poisson regression model to assess medication effects."
 #| echo: FALSE
@@
 #| label: tbl-ma
-#| tbl-cap: "Mediation analysis with estimated effects and 95% CI."
+#| tbl-cap: "Mediation analysis with estimated effects and 95% CI. For negative direct effect or indirect effect, the percentage of mediated effect cannot be calculated due to model setting."
 #| echo: FALSE
@@
 #| label: fig-rf
-#| fig-cap: "Parameter tuning results of random forest. "
+#| fig-cap: "Parameter tuning results of random forest. Darker square means lower root mean squared error."
 #| echo: FALSE
@@
 #| label: fig-rt
-#| fig-cap: "Parameter tuning results of regression tree. "
+#| fig-cap: "Parameter tuning results of regression tree. Darker square means lower root mean squared error. "
 #| echo: FALSE
@@
 #| label: fig-rfrt
-#| fig-cap: "Comparison of random forest and regression tree: predicted values vs. observed values. "
+#| fig-cap: "Comparison of random forest and regression tree: predicted values vs.
+observed values.
+The dashed black line on the diagonal denotes the reference where predicted values are equal to observed values.
+"
 #| echo: FALSE
@@
 #| label: fig-rfsim
-#| fig-cap: "Parameter tuning results of random forest model using simulated data. "
+#| fig-cap: "Parameter tuning results of random forest model using simulated data. Darker square means lower root mean squared error. "
 #| echo: FALSE
@@
 #| label: fig-rfpredsim
-#| fig-cap: "Fitting results of random forest on simulated data: predicted values vs. observed values. "
+#| fig-cap: "Fitting results of random forest on simulated data: predicted values vs.
+observed values.
+The dashed black line on the diagonal denotes the reference where predicted values are equal to observed values.
+"
 #| echo: FALSE
